Designing Types for R, Empirically

### 2020-05-15 paper/main.tex
clenup

\renewcommand{\null}{\texttt{\textbf{null}}\xspace}
\newcommand{\any}{\texttt{\textbf{any}}\xspace}
\newcommand{\environment}{\texttt{\textbf{environment}}\xspace}
\newcommand{\expression}{\texttt{\textbf{expression}}\xspace}
\newcommand{\Language}{\texttt{\textbf{language}}\xspace} %% Upper case!
\newcommand{\externalptr}{\texttt{\textbf{externalptr}}\xspace}
\renewcommand{\symbol}{\texttt{\textbf{symbol}}\xspace}
\newcommand{\pairlist}{\texttt{\textbf{pairlist}}\xspace}
\newcommand{\weakref}{\texttt{\textbf{weakref}}\xspace}
\renewcommand{\int}{\texttt{\textbf{int}}\xspace}
\newcommand{\chr}{\texttt{\textbf{chr}}\xspace}
\newcommand{\dbl}{\texttt{\textbf{dbl}}\xspace}
\newcommand{\lgl}{\texttt{\textbf{lgl}}\xspace}
\newcommand{\clx}{\texttt{\textbf{clx}}\xspace}
\newcommand{\raw}{\texttt{\textbf{raw}}\xspace}
\newcommand{\NA}{\texttt{{NA}}\xspace}
\newcommand{\VARGS}{\texttt{\textbf{...}}\xspace}
\newcommand{\T}{\ensuremath{T}\xspace}
\renewcommand{\S}{\ensuremath{S}\xspace}
\newcommand{\V}{\ensuremath{V}\xspace}
\newcommand{\A}{\ensuremath{A}\xspace}
\newcommand{\ID}{\ensuremath{ID}\xspace}
\newcommand{\F}{\ensuremath{F}\xspace}
\newcommand{\B}{\ensuremath{B}\xspace}
\newcommand{\FUN}[2]{\ensuremath{\langle #1 \rangle \rightarrow #2}\xspace}
\newcommand{\STRUCT}[1]{\texttt{\textbf{struct}}\ensuremath{\langle #1\rangle}\xspace}
\newcommand{\LIST}[1]{\texttt{\textbf{list}}\ensuremath{\langle #1\rangle}\xspace}
\newcommand{\CLASS}[1]{\texttt{\textbf{class}}\ensuremath{\langle #1\rangle}\xspace}
\newcommand{\VEC}[1]{#1\k{[]}\xspace}
\newcommand{\NAVEC}[1]{\k{\string^}\!#1\k{[]}\xspace}

@@

\begin{figure}[!h] \noindent \small  \centering\begin{minipage}{.45\linewidth}
\begin{tabular}{lclr}
\T& ::= & \any          & \it top type\\
  & |   & \null         & \it null type\\
  & |   & \environment  & \it environment type \\
  & |   & \Language     & \it language type \\
  & |   & \symbol       & \it symbol type \\
  & |   & \S            & \it scalar type \\
  & |   & \V            & \it vector type \\
  & |   & \T~\k{|}\T    & \it union type \\
  & |   & \k{?} \T      & \it nullable type \\
  & |   & \FUN{A_1, ..., A_n}\T    & \it function type \\
  & |   & \LIST\T                  & \it list type\\
  & |   & \CLASS{ID_1, ..., ID_n}  & \it class type\vspace{5pt}\\
\end{tabular}\end{minipage}   \hfill
\begin{minipage}{.45\linewidth}\begin{tabular}{lclr}
\A & ::= & \T          & \it arguments \\
   & |   & \VARGS      & \vspace{5pt} \\
\V & ::= & ~\VEC\S     & \\
   & |   & \NAVEC\S    & \it na vector types \vspace{5pt}\\
\S & ::= & \int        & \\
   & |   & \chr \\
   & |   & \dbl \\
   & |   & \lgl \\
   & |   & \clx \\
   & |   & \raw \vspace{5pt}\\
\end{tabular}\end{minipage}\caption{The R type language}\label{types}
\end{figure}

@@

\paragraph{Missing values} In R, a missing value is written
\NA, each of basic types has its specific representation of \NA. Thus there
is an \int missing observation as well as a double missing, and they are
different.  The type system allows to distinguish between vectors that may
contain missing observations (written \NAVEC\int) and those who are
guarantee to be \NA-free (\VEC\int).  A \NA-free vector can be treated as a
vector that may have missing observations.  We do not allow scalar values to
be missings.

>>>>

\renewcommand{\null}{\texttt{\textbf{null}}\xspace}
\newcommand{\any}{\texttt{\textbf{any}}\xspace}
\newcommand{\environment}{\texttt{\textbf{envir}}\xspace}
\newcommand{\expression}{\texttt{\textbf{expression}}\xspace}
\newcommand{\Language}{\texttt{\textbf{lang}}\xspace} %% Upper case!
\newcommand{\externalptr}{\texttt{\textbf{externalptr}}\xspace}
\renewcommand{\symbol}{\texttt{\textbf{symbol}}\xspace}
\newcommand{\pairlist}{\texttt{\textbf{pairlist}}\xspace}
\newcommand{\weakref}{\texttt{\textbf{weakref}}\xspace}
\renewcommand{\int}{\texttt{\textbf{int}}\xspace}
\newcommand{\chr}{\texttt{\textbf{chr}}\xspace}
\newcommand{\dbl}{\texttt{\textbf{dbl}}\xspace}
\newcommand{\lgl}{\texttt{\textbf{lgl}}\xspace}
\newcommand{\clx}{\texttt{\textbf{clx}}\xspace}
\newcommand{\raw}{\texttt{\textbf{raw}}\xspace}
\newcommand{\NA}{\texttt{{NA}}\xspace}
\newcommand{\VARGS}{\texttt{\textbf{...}}\xspace}
\newcommand{\T}{\ensuremath{T}\xspace}
\renewcommand{\S}{\ensuremath{S}\xspace}
\newcommand{\V}{\ensuremath{V}\xspace}
\newcommand{\A}{\ensuremath{A}\xspace}
\newcommand{\ID}{\ensuremath{ID}\xspace}
\newcommand{\F}{\ensuremath{F}\xspace}
\newcommand{\B}{\ensuremath{B}\xspace}
\newcommand{\FUN}[2]{\ensuremath{\langle #1 \rangle \rightarrow #2}\xspace}
\newcommand{\STRUCT}[1]{\texttt{\textbf{struct}}\ensuremath{\langle #1\rangle}\xspace}
\newcommand{\LIST}[1]{\texttt{\textbf{list}}\ensuremath{\langle #1\rangle}\xspace}
\newcommand{\CLASS}[1]{\texttt{\textbf{class}}\ensuremath{\langle #1\rangle}\xspace}
\newcommand{\VEC}[1]{#1\k{[]}\xspace}
\newcommand{\NAVEC}[1]{\k{\string^}\!#1\k{[]}\xspace}

@@

\begin{figure}[!h] \noindent \small  \centering\begin{minipage}{.45\linewidth}
\begin{tabular}{lclr}
\T& ::= & \any          & \it top type\\
  & |   & \null         & \it null type\\
  & |   & \environment  & \it environment type \\
  & |   & \S            & \it scalar type \\
  & |   & \V            & \it vector type \\
  & |   & \T~\k{|}\T    & \it union type \\
  & |   & \k{?} \T      & \it nullable type \\
  & |   & \FUN{A_1, ..., A_n}\T    & \it function type \\
  & |   & \LIST\T                  & \it list type\\
  & |   & \CLASS{ID_1, ..., ID_n}  & \it class type\vspace{5pt}\\
\end{tabular}\end{minipage}   \hfill
\begin{minipage}{.45\linewidth}\begin{tabular}{lclr}
\A & ::= & \T          & \it arguments \\
   & |   & \VARGS      & \vspace{5pt} \\
\V & ::= & ~\VEC\S     & \\
   & |   & \NAVEC\S    & \it na vector types \vspace{5pt}\\
\S & ::= & \int        & \\
   & |   & \chr \\
   & |   & \dbl \\
   & |   & \lgl \\
   & |   & \clx \\
   & |   & \raw \vspace{5pt}\\
\end{tabular}\end{minipage}\caption{The R type language}\label{types}
\end{figure}

@@

\paragraph{Missing values} In R, a missing value is written
\NA, each of basic types has its specific representation of \NA. Thus there
is an \int missing observation as well as a double missing, and they are
different.  The type system allows to distinguish between vectors that may
contain missing observations (written \NAVEC\int) and those who are
guarantee to be \NA-free (\VEC\int).  A \NA-free vector can be treated as a
vector that may have missing observations.  We do not allow scalar values to
be missings.  The type \raw does not allow {\NA}s.
